---
title: "BMI Analysis"
format:
  html:
    code-fold: false
execute:
  echo: false
  warning: false
  messag1e: false
---

```{r}
#| include: false
library(ggplot2)
library(tidyverse)
#df <-read.csv("C:/Users/86185/Downloads/DAProject2.csv",stringsAsFactors = T)
df <-read.csv("DAProject2.csv",stringsAsFactors = TRUE)

```

# Introduction

This project analyses the Body Mass Index (BMI) distributions to investigate how BMI is influenced by a range of socio-economic and lifestyle factors. Using regression-based methods, the analysis aims to quantify the change over the period 2008 to 2012, as well as quantify the relationships between BMI and potential explanatory variables. The overall goal is to assess the strength, significance, and predictive power of the resulting statistical model. Understanding these relationships is important for identifying key factors associated with BMI variation and informing potential public health insights.

## Data Overview

The dataset contains observations collected between 2008 and 2012, including demographic, socio-economic and lifestyle variables such as age, education and sex. The response variable in this analysis is BMI, which provides a quantitative measure for comparing individuals across time. The explanatory variables consist of the aforementioned lifestyle factors, along with age and sex, which are included to assess their potential influence on BMI.

# Exploratory Analysis

```{r}
df <-read.csv("DAProject2.csv",stringsAsFactors = TRUE)
library(tidyverse)
library(skimr)    
df%>%
  select(!Year)%>%
  skim()
```

# ```{r}
# ggplot(df, aes(x = BMI)) +
#   geom_histogram(binwidth = 1 , fill = "coral1", colour = "darkturquoise") +
#   labs(title = "Histogram of BMI scores", xlab = "BMI", ylab = "Frequency")
#   
# ```
# 
# The histogram shows a slightly right skewed distribution, with most BMI values concentrated between approximately 22 and 32. Although this indicates a deviation from normality, the large sample size suggests that linear regression remains appropriate, subject to further diagnostic checks.

```{r}
# Boxplot indicating constant BMI across the years of the study
ggplot(df, aes(x = factor(Year), y = BMI)) +
  
  # Making the plot more readable by reducing it's range on the BMI axis, and excluding outlier points
  geom_boxplot(aes(group = Year), outlier.shape = NA) +
  coord_cartesian(ylim = c(10, 45))
```

```{r}

# Create a useful column for exploratory plots
df$FruitVeg <- interaction(df$Fruit, df$Veg)
# Organise Education level in a more interpretable manner (ranging from lowest to highest level of prestige)
df$Education <- factor(df$Education, 
                            levels = c("No qualifications", "Other school level", 
                                       "Standard grade or equiv", "Higher grade or equiv", 
                                       "HNC/D or equiv", "Degree or higher"))
# Subset of the data which will help to create a more readable exploratory plot
df_subset <- df %>%
  filter(Education %in% c("No qualifications", "Degree or higher"))
# Creates a normalised table indicating the portions of participants at each education level who meet each category of fruit
healthEdu <- table(df$FruitVeg, df$Education)
healthEduNorm <- sweep(healthEdu, 2, colSums(healthEdu), FUN = "/")
rownames(healthEduNorm) <- c("Neither", "Fruit only", "Veg only", "Both")
ggplot(df_subset, aes(x = Age, y = BMI)) + 
  
  geom_point(aes(colour = FruitVeg), size = 1.0) +
  
  facet_grid(Sex ~ Education) +
  
  # Mean line for reference
  stat_summary(
    aes(linetype = "Mean BMI"),       # maps the line to a new legend 
    fun = mean,
    geom = "line",
    colour = "grey20",
    size = 1.0) +
  
  # legend labels
  scale_colour_manual(
    values = c(
      "No.No" = "red",
      "Yes.No" = "orange",
      "No.Yes" = "yellow",
      "Yes.Yes" = "green"),
    
    labels = c(
      "No.No" = "Neither",
      "Yes.No" = "Fruit Only",
      "No.Yes" = "Veg Only",
      "Yes.Yes" = "Both")) +
  
  scale_linetype_manual(
    values = c("Mean BMI" = "solid")) +
  
  # Legend title
  labs(colour = "Recommended fruit/veg intake?", linetype = "")
```

To further explore relationships between lifestyle factors, a chi-squared test was conducted to assess the association between fruit and vegetable consumption. The results of the test indicated whether these variables are independent or exhibit a significant relationship.

```{r}
library(knitr)
chi <- chisq.test(table(df$Fruit, df$Veg))
result <- data.frame(
  Test = "Chi-square test",
  `Chi-square` = round(chi$statistic, 2),
  Df = chi$parameter,
  `P-value` = ifelse(chi$p.value < 0.001, "<0.001", round(chi$p.value, 3))
)
kable(result,
      caption = "Chi-square Test of Association between Fruit and Vegetable Intake",
      booktabs = TRUE,
      align = "c")
```

This was built upon visually using a proportional bar chart, which illustrates the distribution of vegetable consumption across different levels of fruit intake. The plot suggests that there is a strong evidence of association between fruit and vegetable intake between these variables, with proportions varying across categories, indicating that individuals with different fruit consumption levels tend to have differing vegetable consumption patterns.

```{r}
ggplot(df, aes(x = Fruit, fill = Veg)) +
   geom_bar(position = "fill") +
   labs(title='Proportion of Vegetable Intake by Fruit Consumption',
   y = "Proportion")+
  theme_minimal()
```

These findings justify the inclusion of both variables in the regression model, as they may capture related but distinct aspects of dietary behavior.

In addition,quadratic relationship suggests that BMI increases with age up to middle age before gradually decreasing at older ages.

```{r}
 ggplot(df, aes(x = Age, y = BMI)) +
   geom_point(alpha = 0.3) +
   geom_smooth(method = "loess", color = "blue", se = TRUE) +
   labs(
     title = "The relationship between BMI and Age",
     x = "Age",
     y = "BMI") +
   theme_minimal()
```
