---
title: "Formal_analysis2"
format: html
editor_options: 
  chunk_output_type: console

execute: 
  echo: false
  warning: false
  message: false
  eval: true
---

To investigate how BMI varies with demographic and lifestyle factors, a multiple linear regression model was fitted with BMI as the response variable. The explanatory variables included age, sex, education level, fruit intake, and vegetable intake.

Exploratory analysis suggested a non-linear relationship between BMI and age. To account for this, both age and its squared term were included in the model.In addition, selected interaction terms (age by sex, sex by education, and fruit by education) were included to allow for potential differences in effects across subgroups.

```{r}
#| include: false
dataset <- read.csv("DAProject2.csv", stringsAsFactors = TRUE)
#The Full model (Saturated Model) 
full_model <- lm(BMI~Age*Sex*Education*Veg*Fruit*Year, data = dataset)
summary(full_model)
#A more meaningful staurated model would be
meaningful_fm_1 <- lm(BMI~Age + I(Age^2)+Sex + Education + Fruit + Veg + Year + Age*Sex + Age*Education + Sex*Education + Fruit*Veg + Education*Fruit + Education*Veg, data = dataset)
summary(meaningful_fm_1)

aic_1 <- step(meaningful_fm_1, direction = "backward")

```

```{r}
#| include: false
model_initial <- lm(formula = BMI ~ Age + I(Age^2) + Sex + Education + Fruit+
    Veg + Age:Sex + Sex:Education + Fruit:Veg + Fruit:Education, data = dataset)
#Checking if interaction terms are significant 
model_no_interaction <- update(model_initial, . ~ . - Age:Sex)  
anova(model_initial, model_no_interaction) #p-value < 0.05,
model_no_interaction <- update(model_initial, . ~ . - Sex:Education)
anova(model_initial, model_no_interaction) #p-value < 0.05,
model_no_interaction <- update(model_initial, . ~ . - Fruit:Education)
anova(model_initial, model_no_interaction) #p-value < 0.05,
model_no_interaction <- update(model_initial, . ~ . - Fruit:Veg)
anova(model_initial, model_no_interaction)#p-value > 0.05 thus this interaction is not significant
#THE FINAL MODEL FOR THE TWO QUESTION OF INTEREST IS 
model_final <- lm(formula = BMI ~ Age + I(Age^2) + Sex + Education + Fruit+Veg + Age:Sex + Sex:Education  + Fruit:Education, data = dataset)
model_add <- lm(BMI~Age+I(Age^2)+Sex+Education+Veg+Fruit, data = dataset)
anova(model_add, model_final)
#thus model with interaction terms is more significant
```

The final model was chosen to balance explanatory power and interpretability:

$$
BMI_i = \beta_0 + \beta_1 \text{Age}_i + \beta_2 \text{Age}_i^2 + \beta_3 \cdot \mathbb{I}_{\mathrm{Sex}}(Male)+ \sum_{j=1}^{5}\beta_{4j}E_{ij}+\beta_5\cdot \mathbb{I}_{\mathrm{Fruit}}(Yes)+\beta_6\cdot \mathbb{I}_{\mathrm{Veg}}(Yes)+\beta_7 (\text{Age}_i \times \text{Sex}_i)+\sum_{j=1}^{5} \gamma_j (\text{Sex}_i \times E_{ij})+\sum_{j=1}^{5} \delta_j (\text{Fruit}_i \times E_{ij})+\varepsilon_i
$$

------------------------------------------------------------------------

**Question 1**

```{r}
library(tidyverse)
library(ggplot2)
library(performance)
library(see)
library(sjPlot)
library(kableExtra)
library(broom)
library(knitr)
library(car)
```

```{r}
#| eval: false
mean <- dataset%>%
  summarise(mean_BMI = mean(BMI), 
            .by = Year)
ggplot(mean, aes(x = Year, y = mean_BMI))+
  geom_line()
```

From the plot the it is suggested that there is no significant trend followed by the observation over 2008-2012

```{r}
#The Model for question 1 
time_model <- lm(BMI~Year, data = dataset)
summary(time_model)
#we will chech the model diagnostics 
check_model(time_model)


#complementary model and its check no need to include it 
#model_q1 <-lm(formula = BMI ~ Age + I(Age^2) + Education +
    #Veg + Year + Age:Sex + Sex:Education + Fruit:Education, data = dataset)
#Note that I have added the year term in the final model becuase we are checking prevelance with time. 
#summary(model_q1)
#check_model(model_q1)
#check_model(model_q1,check = c("linearity","homogeneity"))
#check_model(model_q1,check = c("qq","normality"))


```

```{r}
tidy(time_model, conf.int = TRUE) %>%
  select(term, estimate, p.value, conf.low, conf.high) %>%
  kable(
    digits = 3, 
    booktabs = TRUE, 
    col.names = c("Term", "Estimate", "P-Value", "Lower CI", "Upper CI") ) %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"), full_width = FALSE, font_size = 8) %>%
  # Adjust column 1 (Predictors) to be 4cm wide
  column_spec(1, width = "4cm") %>%
  # Adjust the rest to be narrower (e.g., 2cm)
  column_spec(2:5, width = "2cm")
```

Here year is not a significant predictor of BMI as the p-value is greater than 0.05, however we may still see an increase in BMI over the years due to the coefficient term being positive. The confidence interval contains 0, further suggesting that there is no prevalence in BMI from year to another.

In conclusion there have been changes in BMI over time as seen from the plot, but changes in year is not what is causing the change in BMI. Thus, variation in BMI is explained by peoples education and lifestyle status rather than the year when data was collected.

------------------------------------------------------------------------

**Question 2**

```{r}
#| include: false
#Fruit and Veg Correlation
##EDA
ggplot(dataset, aes(x = Fruit, fill = Veg)) +
  geom_bar(position = "fill") +
  labs(y = "Proportion") +
  theme_minimal()
contingency_table <- table(dataset$Veg, dataset$Fruit)
print(contingency_table)

##FA_Perform the Chi-squared test
chi_test_result <- chisq.test(contingency_table)
print(chi_test_result)
collinearity <- vif(model_final)
summary(model_final)
```

Based on the relationship of lifestyle factors we found above, even though Fruit and Veg are correlated, the model doesn't reflect high multicollinearity therefore it is reasonable to keep both terms.

```{r}
model_q2 <- model_final
summary(model_q2)
#model diagnostics 
check_model(model_q2,check = c("linearity","homogeneity","qq","normality"))
```

---

**CONFIDENCE INTERVALS** for the beta estimates

```{r}
model_data <- tidy(model_q2, conf.int = TRUE) %>%
  select(term, estimate, p.value, conf.low, conf.high)

model_data %>%
  kbl(booktabs = TRUE, linesep = "", digits = 2,
      col.names = c("Term", "Estimate", "P-Value", "Lower CI", "Upper CI")) %>% 
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"), full_width = FALSE, font_size = 9) %>%
  column_spec(1, width = "4cm") %>%
  column_spec(2:5, width = "2cm") %>%
  row_spec(0:nrow(model_data), extra_css = "padding: 4px;") 
```

\
\
